Simple repression draft

### 2020-05-19 doc/section_04_bayesian_inference.tex
initial splitting of document into several files

\documentclass[12pt]{article}%{amsart}
\usepackage[top = 1.0in, bottom = 1.0in, left = 1.0in, right = 1.0in]{geometry}
% See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}               % ... or a4paper or a5paper or ...
%\geometry{landscape}             % Activate for for rotated page geometry
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
%\usepackage{epstopdf}
% \usepackage{caption}
\usepackage{amsmath}
% \usepackage{longtable}
% \usepackage{tabu}
% \usepackage{accents} %to get undertildes for vec & mat
\usepackage{color} % for colored text
% \usepackage[normalem]{ulem} % for editing: striking out text using 'sout' command

% \newcommand\mytablefigwidth{0.35\textwidth}

% allows to use .ai files directly w/o resaving as pdf
\DeclareGraphicsRule{.ai}{pdf}{.ai}{}

% Handy math macros!
\newcommand{\vect}[1]{\vec{#1}}
\newcommand{\matr}[1]{\mathbf{#1}}
\newcommand{\rate}[3]{{#1}_{#2}^{#3}}
\newcommand{\mmnote}[1]{\textcolor{cyan}{(MM:~#1)}}

% derivative macros. these are sneaky. the [{}] is an empty default 1st arg
% usage: provide 1 arg: \deriv{x} gives d/dx
% provide 2 args like so: \deriv[f]{x} gives df/dx
\newcommand{\deriv}[2][{}]{\frac{d #1}{d #2}}
\newcommand{\pderiv}[2][{}]{\frac{\partial #1}{\partial #2}}
\newcommand{\psecderiv}[2][{}]{\frac{\partial{^2} #1}{\partial #2{^2}}}

%%%%% Referencing macros %%%%%
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\tref}[1]{Table~\ref{#1}}
\newcommand{\eq}[1]{Eq.~(\ref{#1})}
\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}

%%%%%
\begin{document}

\title{Simple repression draft}

\maketitle

\section{Introduction}
\mmnote{Key ideas, no particular order, that I haven't written down before:
\begin{itemize}
\item Our goal is to build phenomenological models of input-output
functions of genetic circuits. More precisely, we want phenomenology
in the sense of coarse-graining away atomistic detail, but still
retaining biophysical meaning, e.g., we don't want to coarse-grain
as far as Hill functions. Why not? We are motivated by studies
like~\cite{Razo-Mejia2020}, for which Hill functions are
insufficient. Such work requires a quantitative theory of how
biophysical changes at the molecular level propagate to
input-output functions at the genetic circuit level.
We want concepts, not mere facts.
In particular a key question is: at this level of
coarse graining, what microscopic details do I need to explicitly
model, and how do we figure that out? For example, do I need to
worry about all or even any of the steps that individual RNAPs go
through each time they make a transcript? Turning the question
around, can we see any imprint of those processes in the
available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling can complement each other
beautifully here.
\end{itemize}}

\fig{fig1:means_cartoons}A shows the qualitative picture of
simple repression. An operator, the binding site on the DNA for a
repressor protein, may be found occupied by a repressor, in which
case transcription is blocked from occurring. Or it may be found
unoccupied, in which case RNA polymerase (RNAP) may bind and
transcription can proceed.

The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
It is often imagined that when the repressor is bound to its operator,
RNAP is sterically blocked from binding to its promoter sequence.
Current evidence suggests this is sometimes, but not
always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
Rob??, showing locations of repressor binding sites far upstream
of promoter}, and it remains an interesting open question precisely
how a repressor bound far upstream is able to repress transription.
Suggestions include ``action-at-a-distance'' mediated by kinks in
the DNA, formed when the repressor is bound, that prevent RNAP binding.
Nevertheless, our modeling in this work is sufficiently
coarse-grained that we simply assume exclusive binding and need
not worry about such precise details.

\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates. In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}

\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple
repression, let us consider what we can learn from theory and
measurements of fold-change in mean expression. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
\label{eq:fc_def}
\end{equation}
where angle brackets denote the averge over a population of cells
and mean mRNA $\langle m\rangle$ is viewed as a function of
repressor copy number $R$.
The second
equality in~\ref{eq:fc_def} follows from assuming that the
translation efficiency, i.e., the number of proteins translated
per mRNA, is the same in both conditions.
In other words, we assume that mean protein level is proportional
to mean mRNA level, and that the proportionality constant is the
same in both conditions and therefore cancels out in the ratio.
This is reasonable since the cells in the two conditions are
identical except for the presence of the transcription factor,
and it is difficult to imagine the transcription factor having
any direct effect on translation.

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
Part of its utility in dissecting transcriptional regulation is
its ratiometric nature, which is always easier than making an
absolute measurement. Also, by measuring otherwise identical cells
with and without a transcription factor present, any biological
noise common to both conditions can be made to cancel away.
In addition, the presence of different operator sites is known to
change the basal expression level in the absence of
repressor~\mmnote{look up cite}.
Computing fold-changes neatly cancels this effect as well.

\fig{fig1:means_cartoons} depicts a smorgasbord
of mathematized cartoons for simple repression that have appeared
in previous literature. For each cartoon, we calculate the
fold-change in mean gene expression as predicted by that model,
deferring some algebraic details to the appendix.
What we will find is that all cartoons collapse to a single
master curve, shown in \fig{fig1:means_cartoons}c, which contains
just two parameters. We label the parameters $\Delta F_R$, an
effective free energy parametrizing the repressor-DNA
interaction, and $\rho$, which subsumes all details of
transcription in the absence of repressors.
We will offer some intuition for why this master curve exists and discuss
why it is likely impossible to discriminate ``right'' from ``wrong''
cartoons given only measurements of fold-changes in expression.

\subsubsection{Equilibrium, 2-state}
In this simplest model, the promoter is idealized as existing in
one of two states, either repressor bound or repressor unbound.
The rate of transcription, and therefore the level of gene
expression, is assumed to be proportional to the fraction of time
spent in the repressor unbound state.
Mathematically, this amounts to the statement that the mean mRNA level is
given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound
state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
copy number, $N_{NS}$ is the number of nonspecific binding sites
in the genome (where repressors spend most of their time when not
bound to the operator, their specifc binding site),
$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
repressor to its operator site. The derivation of this result is deferred
to the appendix.

The fold-change is easily found as the ratio of mean mRNA with
and without repressor. This gives
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
which clearly matches the form of the master curve
in~\eq{fig1:means_cartoons} with $\rho=1$.

In fact it was noted in~\cite{Chure2019} that this two-state
model can be viewed as the coarse-graining of any equilibrium
promoter model in which no transcriptionally active states have
transcription factor bound. Put differently, there is no overlap
between TF bound states and transcriptionally active states. We
will see this explicitly in the 3-state equilibrium model below,
but perhaps surprising is that something analogous carries over
even to the nonequilibrium models we consider later.

\subsubsection{Equilibrium, 3-state}
Compared to the previous model, here we fine-grain the repressor
unbound state into two separate states: empty, and RNAP bound.
This picture was used in~\cite{Garcia2011a} as we use it here,
and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
to incorporate small-molecule inducers that bind the repressor.
Interestingly, the effect of this is simply to renormalize $R$
from the total number of repressors to a smaller effective number,
roughly (though not exactly) the number which are unbound by inducers.

The mean mRNA, as derived in the appendix from a similar
enumeration of states and weights as the previous model, is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
        },
\end{equation}
where the new variables are $\Delta\epsilon_P$, the difference in
RNAP binding energy to its specific site (the promoter) relative
to an average nonspecific background site, and the RNAP copy
number, $P$. The fold-change again follows immediately as
\begin{align}
\text{fold-change}
&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
        }
\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
\\
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
\end{align}
with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
as shown in~\fig{fig1:means_cartoons}.

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{
\textbf{
An overview of the simple repression motif at the level of means.}
(A) depicts the qualitative biological picture of simple repression.
(B) shows a variety of possible mathematized cartoons of simple
repression, along with the effective parameter $\rho$
which subsumes all regulatory details
of the architecture excluding the repressor.
(C) shows the ``master curve'' to which all cartoons in (B) collapse.
}
\label{fig1:means_cartoons}
\end{figure}

\subsubsection{Nonequilibrium model 1 - Poisson promoter}
For our first kinetic model, we consider the simplest possible
picture with only two states, repressor bound and unbound.
This is exactly the model used for the main results of~\cite{Jones2014}.
In this picture, repressor association and dissociation rates
from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
transitions between the two states. When the system is in the
unbound state, transcriptions initiate at rate $r$, which
represents a coarse-graining of all the downstream processes into
a single effective rate. mRNA is again degraded at rate $\gamma$.

Let $p_{m,R}(t)$ denote the joint probability of finding the
system in the repressor bound state with $m$ mRNA molecules
present at time $t$, and similarly define $p_{m,U}(t)$ for the
repressor unbound state. 
This model is governed by a master equation giving the time
evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
\begin{align}
\begin{split}
\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
\\
\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
                        + rp_{m-1,U}(t) - rp_{m,U}(t)
                \\
                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
\label{eq:poisson_promoter_cme}
\end{split}
\end{align}
In each equation, the first two terms describe transitions
between promoter states due to repressors unbinding and binding,
respectively. The final two terms describe degradation of mRNA,
decreasing the copy number by one, and the terms with coefficient
$r$ describe transcription initiation increasing the mRNA copy
number by one.

We can greatly simplify the notation, which will be especially
useful for the more complicated models yet to come, by
re-expressing the master equation in vector form. The promoter
states are collected into a vector and the rate constants are
collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t)
                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix.
Taking steady state by setting time derivatives to zero,
the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to the appendix.
Recall $k_R^+$ is proportional to the repressor copy number, so
in computing fold-change, absence of repressor corresponds to
$k_R^+\rightarrow0$.
Therefore fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.

\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
Analogous to the three-state equilibrium model, this model
fine-grains the repressor unbound state of nonequilibrium model 1,
resolving it into an empty promoter state and an RNAP-bound state.
Note in this picture, in contrast with model 4 below,
transcription initiation is accompanied by a promoter state
change, in keeping with the interpretation as RNAP-bound and
empty states: if an RNAP successfully escapes the promoter and
proceeds to elongation of a transcript, clearly it is no longer
bound at the promoter! Therefore another RNAP must bind before
another transcript can be initiated.

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
The main subtlety arises since transcription initiation
accompanies a promoter state change.
This can be understood by analogy to $\matr{K}$. The
off-diagonals and diagonals of $\matr{K}$ correspond to
transitions arriving at or departing from, respectively, the
promoter state of interest. If transcription initiations are
accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since
the arriving and departing transitions have different initial
copy number of mRNA, unlike for $\matr{K}$ where they are the same.
The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having
$m$ mRNA present along with the promoter having repressor bound,
being empty, or having RNAP bound, respectively.
$\matr{R_A}$ describes probability flux \textit{arriving} at the state
$\vect{p}_m$ from a state with one fewer mRNA, namely
$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
\textit{departing} from the state $\vect{p}_m$
for a state with one more mRNA, namely $\vect{p}_{m+1}$.
$\matr{K}$ is closely analogous to model 1.

Mean mRNA at steady state is found analogously to model 1, with the result
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
\end{equation}
Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
from which we have
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
        \frac{k_P^+ + k_P^- + r}{k_P^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
        \right)^{-1}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
        \right)^{-1},
\end{align}
which follows the master curve with
$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.

\subsubsection{Nonequilibrium model 3
- Multistep transcription initiation and escape}
One might reasonably complain that the first two
``nonequilibrium'' models we have considered are straw men. Their
steady states necessarily satisfy detailed balance which is
equivalent to thermodynamic equilibrium. Why is this the case? At
steady state there is by definition no net probability flux in or
out of each promoter state, but since the promoter states form a
linear chain, there is only one way in or out of the repressor
bound and RNAP bound states, implying each edge must actually
have a net zero probability flux, which is the definition of
detailed balance (usually phrased as equality of forward and
reverse transition fluxes).

Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
the possibility of true nonequilibrium steady-state fluxes
through the promoter states.
We point out that this model was considered previously in~\cite{Mitarai2015}
where a comparison was made with model 1 as used in~\cite{Jones2014}.
The authors of~\cite{Mitarai2015} argued that the additional
complexity is essential to properly account for the noise in the
mRNA distribution. We will weigh in on both models later when we
consider observables beyond fold-change.

The master equation governing this model is
identical in form to model 2 above, namely
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\end{equation}
but obviously with a higher-dimensional state space and different matrices.
The state vector and promoter transition matrix are now
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound,
empty, RNAP closed complex, and RNAP open complex.
Besides increasing dimension by one again, the only new feature in
$\matr{K}$ is the rate $k_O$, representing the (irreversible)
rate of open complex formation from the closed complex.
The initiation matrices are given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to model 2.

The expression for mean mRNA is substantially more complicated
now, as worked out in the appendix where we find
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+ k_O}
        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
\end{equation}
which can be simplified to
\begin{equation}
\langle m\rangle
= \frac{r}{\gamma}
\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
\end{equation}
The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
\begin{align}
\text{fold-change}
&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
\\
&= \left(
        1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
\right)^{-1},
\end{align}
surprisingly reducing to the master curve once again, with
$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.

This example hints that an arbitrarily fine-grained model of
downstream transcription steps may still be collapsed to the form
of the master curve in~\fig{fig1:means_cartoons}B so long as the
repressor binding is exclusive with transcriptionally active
states. We offer this as a conjecture, and we suspect that a
careful argument using the King-Altman diagram
method~\cite{King1956, Hill1966} might furnish a ``proof.''
Our focus here is not on full generality but rather to survey an
assortment of plausible models that have been proposed in the
literature for simple repression.

\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
This model is at the core of the theory in~\cite{Razo-Mejia2020}.
At a glance the cartoon for this model may appear very similar to model 2,
and mathematically it is, but the interpretation is rather different.
In model 2, we interpreted the third state literally as an
RNAP-bound promoter and modeled initiation of a transcript as
triggering a promoter state change, making the hopefully
uncontroversial assumption that an RNAP can only make one
transcript at a time. In contrast, in the present model the
promoter state does \textit{not} change when a transcript is
initiated. So we no longer interpret these states as literally
RNAP bound and unbound but instead as coarse-grained ``active''
and ``inactive'' states, the details of which we leave
unspecified for now. We will comment more on this model below
when we discuss Fano factors of models.

Mathematically this model is very similar to models 1 and 2.
Like model 1, the matrix $R$ describing transcription initiation
is diagonal, namely
\begin{equation}
\matr{R} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
        \end{pmatrix}.
\end{equation}
so the master equation takes verbatim the same form as it did for model
1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
reiterate that these are not simply RNAP binding and unbinding rates.

Carrying out the algebra, the mean mRNA can be found to be
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
\frac{k_R^- k^+}
{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
\end{equation}
and the fold-change readily follows,
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \left(1 + \frac{k^+}{k^-}\right)^{-1}
        \right)^{-1},
\end{align}
from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.

\subsubsection{Nonequilibrium model 5 - Bursty promoter}
We will have much more to say about this model in subsequent sections.
For now we sketch it as an intuitive analog to model 1,
with just two states, repressor bound or unbound, and transition rates
between them of $k_R^+$ and $k_R^-$.
In model 1, when in the unbound state, single mRNA transcripts
are produced as a Poisson process with some characteristic rate
$r$. This model by contrast produces, at some Poisson rate $k_i$,
\textit{bursts} of transcripts with a geometrically distributed
burst size and a mean burst size $b$.

From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
\begin{equation}
\langle m\rangle = \frac{k_i b}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
which will turn out to be correct from a careful calculation.
For now, we simply note that just like model 1, the fold-change becomes
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
\end{equation}
with $\rho=1$ also like model 1.

We will also see later how this model emerges as a natural limit of model 4.

\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
What have we learned from our exploration of the models
in~\fig{fig1:means_cartoons} and their predictions of
fold-changes in mean expression levels? The takeaway is the
existence of a master curve (\fig{fig1:means_cartoons}C) to which
the fold-change predictions of all the models collapse. This
master curve is parametrized by only two effective parameters:
$\Delta F_R$, which characterizes the repressor, and $\rho$,
which characterizes all other features of the promoter
architecture. The key assumption underpinning this result is that
zero transcription occurs when a repressor is bound to its
operator. Note, however, that we are agnostic about the molecular
mechanism which achieves this;
steric effects are the obvious mechanism, but, for instance,
``action-at-a-distance'' mediated by kinked DNA due to
repressors bound tens or hundreds of nucleotides upstream of a
promoter is plausible as well.

Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
This brings to mind the deep questions posed in,
e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
for general ``symmetries'' of the problem instead of specifying
generative models of the known microscopics of a system.
In our case, the key feature seems to be the exclusive nature of
repressor and RNAP binding, which allows the parameter describing
the repressor, $\Delta F_R$, to cleanly separate from all other
details of the promoter architecture, which are encapsulated in $\rho$.
In the graph-theoretic language which was applied to similar
problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
repressor-bound state is a leaf of the graph describing
transitions between promoter states. Arbitrary nonequilibrium
behavior can occur on the rest of the promoter state space, but
it may all be swept up in the effective parameter $\rho$, to
which the repressor makes no contribution.

As suggested in~\cite{Chure2019}, we believe this master curve
should generalize to architectures with multiple repressor
binding sites, as long as the exclusivity of transcription factor
binding and transcription initation is maintained. The
interpretation of $\Delta F_R$ is then of an effective free
energy of all repressor bound states.
In an equilibrium picture this is simply given by the log of the
sum of Boltzmann weights of all repressor bound states, which
looks like the log of a partition function of a subsystem.
In a nonequilibrium picture, while we can still mathematically gather
terms and give the resulting collection the label $\Delta F_R$,
it is unclear if the physical interpretation as an effective free energy
makes sense.
As is well known, free energies cannot be assigned unambiguously
to states out of equilibrium because the free energy change along
a generic path traversing the state space is path dependent,
unlike at equilibrium.
A consequence of this is that, out of equilibrium, $\Delta F_R$
is no longer a simple sum of Boltzmann weights. Instead it
resembles a restricted sum of King-Altman
diagrams~\cite{King1956, Hill1966}. Following the work of
Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
as an effective free energy, but this remains unclear to us.
We leave this an open problem for future work.

If we relax the requirement of exclusive repressor-RNAP binding,
one could imagine models in which repressor and RNAP doubly-bound
states are allowed, where the repressor's effect is to
\text{reduce} the transcription rate rather than setting it to zero.
Our results do not strictly apply to such a model, although we
note that if the repressor's reduction of the transcription rate
is substantial, such a model might still be well-approximated by
one of the models in~\fig{fig1:means_cartoons}.

One may worry that our ``one curve to rule them all'' is a
mathematical tautology. In fact we \textit{agree} with this
criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
be meaningfully interpreted as a real, physical free energy.
An analogy to Hill functions is appropriate here. One of their
great strengths and weaknesses, depending on the use they are put
to, is that their parameters coarse-grain many details and are
generally not interpretable in terms of microscopic models, for
deep reasons discussed at length in~\cite{Frank2013}.
By contrast, our master curve claims to have the best of both worlds:
a coarse-graining of all details besides the repressor into
a single effective parameter $\rho$,
while simultaneously retaining an interpretation of
$\Delta F_R$ as a physically
meaningful and interpretable free energy.
Our task, then, is to prove or disprove this claim.

How do we test this and probe the theory with fold-change
measurements? There is a fundamental limitation in that the
master curve is essentially a one-parameter function of $\Delta
F_R + \log\rho$. Worse, there are many \textit{a priori}
plausible microscopic mechanisms that could contribute to the
value of $\rho$, such as RNAP binding and escape
kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
RNAP clusters analogous to those obvserved in
eukaryotes~\cite{Cisse2013, Cho2016}
and recently also observed in bacteria~\cite{Ladouceur2020}.
Even if $\Delta F_R$ is measured to
high precision, inferring the potential microscopic contributions
to $\rho$, buried inside a log no less, from fold-change
measurements seems hopeless. As a statistical inference problem
it is entirely nonidentifiable, in the language
of~\cite{Gelman2013}, Section 4.3.

If we cannot simply infer values of $\rho$ from measurements of
fold-change, can we perturb some of the parameters that make up
$\rho$ and measure the change? Unfortunately we suspect this is
off-limits experimentally: most of the potential contributors to
$\rho$ are global processes that affect many or all genes. For
instance, changing RNAP association rates by changing RNAP copy
numbers, or changing supercoiling kinetics by changing
topoisomerase copy numbers, would massively perturb the entire
cell's physiology and confound any determination of $\rho$.

One might instead imagine a bottom-up modeling approach, where we
mathematize a model of what we hypothesize the important steps
are and are not, use in vitro data for the steps deemed
important, and \textit{predict} what $\rho$ should be.
But again, because of the one-parameter nature of the master
curve, many different models will likely make indistinguishable
predictions, and without any way to experimentally perturb in
vivo, there is no clear way to test whether the modeling
assumptions are correct.

In light of this, we prefer the view that parameters and rates
are not directly comparable between cartoons
in~\fig{fig1:means_cartoons}. Rather, parameters in the
simpler cartoons represent coarse-grained combinations of
parameters in the finer-grained models. For instance, by equating
$\rho$ between any two models, one can derive various possible
correspondences between the two models' parameters. Note that
these correspondences are clearly not unique, since many possible
associations could be made. It then is a choice as to what microscopic
interpretations the model-builder prefers for the parameters in a
particular cartoon, and as to which coarse-grainings lend
intuition and which seem nonsensical.
Indeed, since it remains an open question what microscopic
features dominate $\rho$ (as suggested above, perhaps RNAP
binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
or, something more exotic like
RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
we are hesitant to put too much weight on any one microscopic
interpretation of model parameters that make up $\rho$.

One possible tuning knob to probe $\rho$ that would not globally
perturb the cell's physiology is to manipulate RNAP binding sites.
Work such as~\cite{Kinney2010} has shown that models of
sequence-dependent RNAP can be inferred from data, and the
authors of~\cite{Brewster2012} showed that the model
of~\cite{Kinney2010} has predictive power by using the model to
\textit{design} binding sites of a desired affinity.
But for our purposes, this begs the question: the authors
of~\cite{Kinney2010} \textit{assumed} a particular model
(essentially our 3-state equilibrium model but without the
repressor), so it is unclear how or if such methods can be turned
around to \textit{compare} different models of promoter function.

>>>>

% !TEX root = ./busty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference and model selection}
\subsection{Parameter inference for constitutive promoters}
